Verifying Spatial Properties of Stencil Computations

### 2016-11-16 paper.tex
Added commented-out paragraph about number of region operators in multi-action stencil specs. (You decide whether to uncomment).
various edits

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of \emph{linearity} in linear type
systems~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

Our terminology of \emph{index schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as index schemes do not bind names.

@@

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note, $\infty$ for multiplication above, \eg{} $2
 \infty = \infty$.
\end{example}
\noindent
The final two equations of \Cref{fig:region-model} give the models
of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
 union of the models of subterms. The model of
$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
more involved. We give some informal intuition first.

@@

\begin{figure}
\vspace{-1em}
\input{spatial-model.tex}
\caption{%\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-0.75em}
\end{figure}

@@

\begin{theorem}[Soundness of equational theory]
\begin{equation*}
\forall S, T, n . \quad
S \equiv T \; \Rightarrow \;
\interp{S}_n = \interp{T}_n
\end{equation*}
(See the supplement for the proof)
\end{theorem}

@@

\begin{theorem}[Approximation soundness]
Code consistent with a specification $S$
 is consistent with a more general specification $T$,
\ie{}, for all models $M$ of a piece of code:
\begin{equation*}
S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
\Rightarrow \; (\consA{M}{\interp{T}_n})
\end{equation*}
where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
(See the supplement for the proof)
\end{theorem}

@@

\begin{example}
\label{exm:checking}
We demonstrate the analysis and modelise procedures over the
following example program which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
  do j = 1, (m-1)
    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
    b(i, j) = (a(i, j) + x + y) / 5.0
end do; end do
\end{minted}
%
The \textit{analyse} function matches on
line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y}) = $
%
%\begin{align*}
{{ %{\small{
$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
%\end{align*}
Thus, \emph{modelise} returns the following
as all multiplicities are one:
%%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%%
\end{example}

@@

\begin{example}
  Consider the following five-point stencil kernel with a (quite general) specification:
%
\begin{minted}[breakindent=8.5em]{fortran}
!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to line three as $l3$ for brevity,
which is checked via
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
with steps:
\begin{enumerate}
%
\item Generate the model from the specification as in
  \cref{sec:semantics}.
%
\vspace{-0.4em}
\begin{align*}
 N_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (N_s))
\end{align*}
\vspace{-1.5em}

\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
  \textsf{once} (M_c)$ where:
%
\vspace{-0.7em}
\begin{align*}
M_c = & \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\end{align*}
\vspace{-1em}
%
\item By the definition of consistency (\Cref{fig:consistency})
 $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
  $\textsf{once}(M_c)$ when:
  \begin{align*}
\vspace{-0.3em}
    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
\vspace{-0.3em}
  \end{align*}
%
\item This holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in N_s . u \preceq_2 v$:
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
   \hspace{-1.5em}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  %
  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
  \preceq_n v$:
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
\end{enumerate}
Since the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{example}
\begin{example}
Consider an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
%
%\begin{minted}[firstnumber=3]{fortran}
%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
%\end{minted}

@@

Abstract syntax trees of the specification language
are synthesised from the set $\spanOp(M)$ of coverings spans.
This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}) and which is wrapped
in the $\textsf{Approx}$ type (see
\Cref{def:mult-and-approx}).  The algebra mirrors the shape
of the region operators $\term{+}$ and $\term{*}$ 
with binary operators $\oplus$ and $\odot$ on
$\mathsf{Approx}(\specDNF)$.
A function $\mathsf{toSpec}$ maps a dimension identifier
and a pair of lower and upper bound
values (one-dimensional) to a specification with a one-dimensional region.
%%
\begin{align*}
\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
\end{align*}
%%
An entire span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
specifications by $\odot$. The resulting specifications for
each span are then summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) =
\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
We unpack the definitions of this algebra.

@@

\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the \trule{\textsc{dist}} rule of the equational theory. The operation is
commutative and associative, and sound with respect to our
model (see \Cref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
where in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
as $\odot$ is commutative.
\end{definition}

\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
$\specDNF$:
%%
\begin{equation*}
S \,\hat{\oplus}\, \epsilon = S \quad
\epsilon \,\hat{\oplus}\, S = S \quad
S \, \hat{\oplus} \, T = S \, \term{+} \, T
\end{equation*}
%%
Thus, if neither specification is empty, their $\hat{\oplus}$ is just
the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in the previous
definition.
\end{definition}

\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
$\forall S,
  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus, our model validates specification synthesis.
%
\subsection{Step 5: Simplification}
The final step applies a simplification pass, orienting the
equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
based on some simple heuristics.

For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
\end{minted}
%%

@@

We built a corpus of around 1 million lines of Fortran code from a
range of scientific computing packages: The Unified Model~\cite{um},
E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
on \cite{griebel1997numerical}), Computational Physics
2~\cite{nicholas2006computational},
ARPACK-NG~\cite{arpackng}, and
SPECFEM3D~\cite{specfem3d}.

We first examined how frequently stencil computations occur. We parsed
959,427 lines of Fortran code and found that 10\% (97,439) of
statements have a left-hand side is an array subscript on
neighbourhood indices. This supports the idea that stencil-like
computations are common in scientific code.

We used CamFort in inference mode to identify stencils within the
corpus. This assumes that the programs have been implemented
correctly---normally one would use CamFort in verification mode to
check the correctness of new code.

We would not expect to infer a stencil for each array statement we
found because our stencil language places significant restrictions on
the types of array-subscript-statements that we will class as a
stencil. In fact, CamFort was able to infer a stencil from 30\% of
these statements. A single statement can produce multiple stencils and
we ended up with 60,525 stencils. This shows that we can still express
a large number of stencil uses within our high-level abstraction.

@@

@@

Various deductive verification tools can express array indexing
in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
for C (see \eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
can be given for a stencil computation, 
however, such specification would replicate the fine-grained
indexing seen in the code and would be similarly prone to indexing
errors. Our approach is much more abstract-- it does
not aim to reify indexing in the specification, but
provides simple spatial descriptions which capture a large number
of common patterns.

@@

% GPGPU
Our work has some similarities with efforts to verify kernels
written for General-Purpose Graphics Processing Unit (GPGPU)
programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
%kernel, and GPGPU programming can be viewed as a massively parallel
%method of transforming a large matrix. 
However, their focus is
mainly on the synchronisation of kernels and the avoidance of
data races, while we are interested in correctness 
embedded within a more typical general-purpose programming
language. %Other work on GPGPU computation, such as
%\citet{Zhang:2012:CGO}, has focused primarily on generating
%optimised code based on relatively simple specifications: to
%provide performance while keeping the programmer's effort within
%reason.

% Sketching Stencils
\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
unoptimised ``reference'' stencils, coupled with partial
implementations which are then completed by a code
generation tool. %These kinds of specifications are just simple
%implementations, so this tool is useful for hand-written, optimised
%stencils. 
The primary purpose of this tool is optimisation rather than
correctness, and the language of specification is much different and
more elaborate than in our approach.

@@

Our definition of a stencil computation meant that we generated no
specification for a large number of iterated array accesses in the
corpus, such as stencils that incorporates reductions. For
future work, we will be looking into expanding the flexibility of 
our analysis to include these examples.

>>>>

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

%Our terminology of \emph{index schemes} and
%\emph{generalisation} is by analogy
%with the \emph{type schemes} and \emph{type generalisation} of
%polymorphic types in ML~\cite{milner1978theory}. Our notion is
%much more rigid however, as index schemes do not bind names.

@@

\begin{example}
For a $2$-dimensional stencil computation, then
\begin{align*}
& \interp{\stenCenS{2}{1}}_2 \\[-0.1em]
& \qquad = \; \{i\singleEntry{1}{2} \,|\, i \in \{\text{-}2,..., 2\}\!\setminus\!0 \}
  \cup  \{ \zeroEntry{1}{2} \,|\, \textit{true} \} \\[-0.2em]
& \qquad = \; \{-2\singleEntry{1}{2}, \, -\singleEntry{1}{2}, \,
  \singleEntry{1}{2}, \, 2\singleEntry{1}{2}\} \cup
  \{\zeroEntry{1}{2}\} \\[-0.2em]
& \qquad = \; \{\vtwoh{-2}{\infty}, \vtwoh{-1}{\infty}, \vtwoh{0}{\infty},
\vtwoh{1}{\infty}, \vtwoh{2}{\infty}\}
\end{align*}
Note, $\infty$ is absorbing for multiplication above \eg{} $2
 \infty = \infty$.
\end{example}
\noindent
The final two equations of \Cref{fig:region-model} give the models
of $\term{+}$ and $\term{*}$. For $\term{+}$ this is the
 union of the models of subterms. The model of
$\term{*}$ takes a ``tensor'' $\otimes$ of subterm models. The tensor is
more involved. We give some informal intuition first.

@@

\begin{figure}
\vspace{-1em}
\input{spatial-model.tex}
\caption{%\textit{spec} model,
$\interp{-}_n : \textit{spec} \rightarrow \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$}
\label{fig:spatial-model}
\vspace{-1em}
\end{figure}

@@

\begin{theorem}[Soundness of equational theory]
\begin{equation*}
\forall S, T, n . \quad
S \equiv T \; \Rightarrow \;
\interp{S}_n = \interp{T}_n
\end{equation*}
See the supplement for the proof.
\end{theorem}

@@

\begin{theorem}[Approximation soundness]
Code consistent with a specification $S$
 is consistent with a more general specification $T$,
\ie{}, for all models $M$ of a piece of code:
\begin{equation*}
S <: T \; \wedge (\consA{M}{\interp{S}_n}) \;
\Rightarrow \; (\consA{M}{\interp{T}_n})
\end{equation*}
where $\consAName$ implements consistency in checking (\cref{subsec:checking}).
See the supplement for the proof.
\end{theorem}

@@

\begin{example}
\label{exm:checking}
We demonstrate the analysis and modelise procedures over the
following example kernel which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
b(i, j) = (a(i, j) + x + y) / 5.0
\end{minted}
%
%The \textit{analyse} function matches on
%line 4 with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
%  y}) = $
%
%\begin{align*}
%{{ %{\small{
%$\{$\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
%  \texttt{a(i, j+1)}, \texttt{a(i, j)}$\}$}}. 
%\end{align*}
The  multiplicities are all one, and so \emph{modelise} yields:
%%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%%
\end{example}

@@

\begin{example}
  Consider the following five-point stencil kernel with a (quite general) specification:
%
\begin{minted}[breakindent=8.5em]{fortran}
!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to line three as $l3$ for brevity,
which is checked via
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$
with steps:
\begin{enumerate}
%
\item Generate the model from the specification as in
  \cref{sec:semantics}.
%
\vspace{-0.4em}
\begin{align*}
 N_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (N_s))
\end{align*}
%\vspace{-1.6em}

\item Generate code model $\textit{modelise}(l3, \texttt{b}) =
  \textsf{once} (M_c)$ where:
%
\vspace{-0.9em}
\begin{align*}
M_c = & \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\end{align*}
\vspace{-1.1em}

\item By the definition of consistency (\Cref{fig:consistency})
 $\textsf{once}(\textsf{exact}(N_s))$ is consistent with
  $\textsf{once}(M_c)$ when:
\vspace{-0.6em}
  \begin{align*}
    & \consSub{2}{\textsf{up}(N_s)}{M_c} \;\,
    \wedge \;\, \consSub{2}{\textsf{low}(N_s)}{M_c}
  \end{align*}
%\vspace{-1.6em}

\item This holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in N_s . u \preceq_2 v$:
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
   \hspace{-1.5em}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
        \hspace{-1.5em}\vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  %
  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
  \preceq_n v$:
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \hspace{-1.3em}\vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \hspace{-1.3em}\vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
\end{enumerate}
Since the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{example}
\begin{example}
Consider an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)}, call it $l3'$.% such that $l3'$ is:
%
%\begin{minted}[firstnumber=3]{fortran}
%a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
%\end{minted}

@@

Abstract syntax trees of the specification language
are synthesised from the set $\spanOp(M)$ of coverings spans.
This conversion is defined in terms
of an algebra on \textit{spec} syntax in disjunctive-normal form
(which we denote \specDNF{}) and which is wrapped
in the $\textsf{Approx}$ type (see
\Cref{def:mult-and-approx}).  The algebra mirrors the shape
of the region operators $\term{+}$ and $\term{*}$ 
with binary operators $\oplus$ and $\odot$ on
$\mathsf{Approx}(\specDNF)$.
A function $\mathsf{toSpec}$ maps a dimension identifier
and a pair of lower and upper bound
values (one-dimensional) to a specification with a one-dimensional region.
%%
\begin{align*}
\mathsf{toSpec} & : \mathbb{N}_{>0} \rightarrow (\domainVal{} \times \domainVal{}) \rightarrow \mathsf{Approx}(\specDNF)
\end{align*}
%%
A span is converted into an $n$-dimensional
specification by applying $\mathsf{toSpec}$ to each
pair of lower and upper bounds in a span, per dimension, and multiplying these
by $\odot$. The resulting specifications for
each span are summed by $\oplus$:
\begin{equation*}
\textsf{spec}(M) =
\oplusbig\limits_{\vect{x} \in \spanOp(M)} \,
\odobig\limits_{i \in \{1, \ldots, n\}} \, \mathsf{toSpec} \; i \; (\vect{x}^L_i, \vect{x}^U_i)
\end{equation*}
We unpack the definitions of this algebra.

@@

\begin{definition} The $\odot$ operation on $\textsf{Approx}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the \trule{\textsc{dist}} rule of the equational theory. The operation is
commutative and associative, and sound with respect to our
model (see \Cref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$ in \Cref{fig:odotlift} where
in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
as $\odot$ is commutative.
\end{definition}

\begin{figure}[t]
\vspace{-1em}
{\small{
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
\vspace{-2em}
}}
\caption{Lifting $\hat{\odot}$ to approximations.}
\label{fig:odotlift}
\end{figure}

\begin{definition}The $\oplus$ operation on $\textsf{Approx}(\specDNF)$
is defined similarly to $\odot$, with the intermediate $\hat{\oplus}$ on
$\specDNF$:
%%
\begin{equation*}
S \,\hat{\oplus}\, \epsilon = S \quad
\epsilon \,\hat{\oplus}\, S = S \quad
S \, \hat{\oplus} \, T = S \, \term{+} \, T
\end{equation*}
%%
Thus, if neither specification is empty, their $\hat{\oplus}$ is just
the syntactic sum $\term{+}$. Then $\oplus$ is the lifting of
$\hat{\oplus}$ to \textsf{Approx} with the shape as $\odot$ in Figure~\ref{fig:odotlift}.
\end{definition}

\begin{lemma}%[$\hat{\odot}$, $\hat{\oplus}$ soundness]
$\forall S,
  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus, our model validates specification synthesis.
%
\subsection{Step 5: Simplification}
The final step applies a simplification pass, orienting the
equational theory $\equiv$ (\Cref{sec:eqs}) into rewrite rules
based on some simple heuristics.
For our example, we infer and then synthesise the
final 5-point specification:
%%
\begin{minted}[linenos=false,breakindent=2em]{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2) + centered(depth=1, dim=2)*pointed(dim=1):: a
\end{minted}
%%

@@

To study the effectiveness of our approach, 
we built a corpus of around 1 million lines of Fortran code from a
range of scientific computing packages: The Unified Model~\cite{um},
E3MG~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
on \cite{griebel1997numerical}), Computational Physics
2~\cite{nicholas2006computational},
ARPACK-NG~\cite{arpackng}, and
SPECFEM3D~\cite{specfem3d}.

We first examined how frequently stencil computations occur. We parsed
959,427 lines of Fortran code and found that 10\% (97,439) of
statements have a left-hand side as an array subscript on
neighbourhood indices. This supports the idea that stencil-like
computations are common in scientific code.

To used the inference procedure of the previous section to identify
stencils within the corpus. This assumes that the programs have been implemented
correctly---normally one would use our approach in verification mode to
check the correctness of new code.

We would not expect to infer a stencil for each array statement we
found because our specification restricts 
the array-subscript-statements that we class as a
stencil. In fact, we was able to infer a stencil from 30\% of
these statements. A single statement can produce multiple stencils and
we ended up with 60,525 stencils. This shows that we can still express
a large number of stencil uses within our high-level abstraction.

@@

% multiActionRegionOps1: 31
% multiActionRegionOps2: 115
% multiActionRegionOps8: 1
% multiActionRegionOps14: 5
% multiActionRegionOps17: 4

% %%%%
% Complexity-wise, out of these 156 multi-action specifications, 31 of
% them used a single region operator, 115 used two region operators, 1
% used eight region operators, 5 used fourteen region operators, and 4
% used seventeen region operators. Some of these very complex stencils
% were found in the atmospheric models of the UM.
% %%%%

@@

Various deductive verification tools can express array indexing
in their specifications, \eg{}, ACSL of \citet{baudin2008acsl}
for C (\eg{}~\citet[Example 3.4.1]{burghardt2010acsl}). A specification
can be given for a stencil computation but replicating the fine-grained
indexing seen in the code and therefore similarly prone to indexing
errors. Our approach is much more abstract-- it does
not aim to reify indexing in the specification, but
provides simple spatial descriptions which capture a large number
of common patterns.

@@

% GPGPU
Our work has similarities with efforts to verify kernels
written for General-Purpose GPU
programming, such as in \citet{Blom:2014:SoCP}. %Stencils are a form of
%kernel, and GPGPU programming can be viewed as a massively parallel
%method of transforming a large matrix. 
However, their focus is
mainly on the synchronisation of kernels and the avoidance of
data races, while we are interested in correctness 
embedded within a more typical general-purpose programming
language. %Other work on GPGPU computation, such as
%\citet{Zhang:2012:CGO}, has focused primarily on generating
%optimised code based on relatively simple specifications: to
%provide performance while keeping the programmer's effort within
%reason.
%
% Sketching Stencils
\citet{Solar-Lezama:2007:PLDI} give specifications of stencils using
unoptimised ``reference'' stencils, coupled with partial
implementations which are completed by a code
generation tool. %These kinds of specifications are just simple
%implementations, so this tool is useful for hand-written, optimised
%stencils. 
The primary purpose of this tool is optimisation rather than
correctness, and the language of specification is much different and
more elaborate than in our approach.

@@

Our restrictive definition of stencils means we do note generate
specifications for a large number of iterated array accesses in the
corpus, \eg{} that incorporate reductions. For
future work, we will look at expanding the flexibility of 
our analysis to include these examples.
